Temporality of the Victorian Hospital Patient Transfer Network

### 2025-08-05 reports/paper.qmd
some progress, mostly power law iftting

```{python}
#| label: compute-imports
import numpy as np
from pathlib import Path
import polars as pl
from scipy import stats
from matplotlib import pyplot as plt, dates as mdates, colors as mcolors, cm
import seaborn as sns
import calendar
from itertools import accumulate
from collections import defaultdict
import datetime
import copy

@@

```{python}
def gof_exponential(bin_edges, counts):
  bin_midpoints = (bin_edges[1:] + bin_edges[:-1]) / 2
  sample_mean = np.sum(bin_midpoints * counts) / np.sum(counts)
  lambd = 1./ sample_mean

  cumul_f = np.exp(-lambd * bin_edges)
  prop_f = -np.diff(cumul_f) * np.sum(counts)

  test_result = stats.chisquare(
    f_obs=counts,
    f_exp=prop_f,
    ddof=len(counts) - 2,
  )

  return test_result, lambd

```

@@

fig, ax = plt.subplots()
sns.scatterplot(stay_dur_df, x='bin_edge_right', y='count', ax=ax)
ax.set_xlabel('Length of Stay [days]')
ax.set_ylabel('Count')
ax.set_xscale('log')
ax.set_yscale('log')

@@

>>>>

```{python}
#| label: compute-imports
import numpy as np
from pathlib import Path
import polars as pl
from scipy import stats, special, signal, optimize
from matplotlib import pyplot as plt, dates as mdates, colors as mcolors, cm
import seaborn as sns
import calendar
from itertools import accumulate
from collections import defaultdict
import datetime
import copy

@@

```{python}
# power law fitting utils
def pull_xfx(df, make_midpoint=True, left="bin_edge_right", right="bin_edge_right", quantity='bin_midpoint', count="count"):
  if make_midpoint:
    xfx_df = df.with_columns(bin_midpoint=(pl.col(left)+pl.col(right))/2)
  else:
    xfx_df = df
  x_arr = xfx_df.select(quantity).to_series().to_numpy()
  y_arr = xfx_df.select(count).to_series().to_numpy()
  return x_arr, y_arr

def estimate_alpha(x, fx, xmin, discr_corr=0.5):
  mask = x > xmin
  relative_x = np.log(x[mask] / (xmin-discr_corr))
  sum_relx = fx[mask] * relative_x
  return 1 + np.sum(fx[mask]) / np.sum(sum_relx)

def scaled_N(x, fx, xmin):
  mask = x > xmin
  return np.sum(fx[mask])

def power_law_density_fn(xs, alpha, xmin, N=1):
  return N * (xs ** (-alpha) / special.zeta(alpha, xmin))

# expon fitting utils
def estimate_gamma(x, fx):
  mean_val = np.sum(x*fx) / np.sum(fx)
  return 1./ mean_val

def expon_density_fn(xs, gamma, N=1):
  return N * gamma * np.exp(-gamma * xs)

# power law graphical method
def power_law_fit_graphical(x, fx, xmin):
  mask = (fx > 0) & (x > xmin)
  coefs = stats.linregress(np.log(x[mask]), np.log(fx[mask]))
  return coefs.slope, np.exp(coefs.intercept)
```

@@

stay_dur_xfx = pull_xfx(stay_dur_df)
stay_dur_expon_gamma = estimate_gamma(*stay_dur_xfx)
stay_dur_pow_alpha = estimate_alpha(*stay_dur_xfx, xmin=40)

fig, ax = plt.subplots()
sns.scatterplot(stay_dur_df, x='bin_edge_right', y='count', linewidth=0, ax=ax)
ax.set_xlabel('Length of Stay [days]')
ax.set_ylabel('Count')
ax.set_xscale('log')
ax.set_yscale('log')

@@

## Fitting power-law distributions for binned data

Due to the underlying data being identiiable health information, we extract binned data for analysing the distributions of patient admission durations and inter-admission duration. We also right truncate the inter-admission duration [check]

For the distribution of patient stay durations, we use the method from @clauset_power-law_2009, where the probability distribution function is given by

$$
f(x) = \frac{\alpha - 1}{x_{\textrm{min}}} \left(\frac{x}{x_{\textrm{min}}\right)^\alpha,
$$

where $\alpha and x_{\textrm{min}}$ are distributional parameters. For a given $x_\textrm{min}$, the MLE for $\alpha$ is recovered as

$$
\hat{\alpha}(x_\text{min}) = 1 + n \left[\sum_{i=1}^{n} \log\left(\frac{x_i}{x_\textrm{min}}\right) \right].
$$

$x_\text{min}$ is estimated by a grid search, and looking the region where the corresponding $\hat{\alpha}\text{s}$ are stable.

```{python}
#| label: fig-stay-dur-xmin-gridsearch
#| fig-cap: Stay duration grid search 
stay_dur_df = pl.read_csv(root / "data/export/length_of_stay_binned.csv")

stay_dur_xfx = pull_xfx(stay_dur_df)

xmin_arr = np.arange(1, 400)
alpha_hats = [estimate_alpha(*stay_dur_xfx, xmin=xmin) for xmin in xmin_arr]
Ncounts = [scaled_N(*stay_dur_xfx, xmin=xmin) for xmin in xmin_arr]

fig, axs = plt.subplots(ncols=2, figsize=(10, 4))
axz = axs.flatten()

axz[1].loglog(xmin_arr, alpha_hats)
axz[1].set_xlabel(r'$x_\text{min}$')
axz[1].set_ylabel(r'$\hat\alpha$')

axz[0].scatter(*stay_dur_xfx, alpha=0.3, color='k', zorder=999)
axz[0].set_xscale('log')
axz[0].set_yscale('log')
axz[0].set_xlabel('Stay Duration [Days]')
axz[0].set_ylabel('Count')
ylim_ = axz[0].get_ylim()

xgrid = np.linspace(0.5, 2500, 10_000)
xgrid_norm = mcolors.Normalize(vmin=min(xmin_arr), vmax=max(xmin_arr))
xgrid_clrs = cm.ScalarMappable(cmap=cm.YlOrRd, norm=xgrid_norm)
for xmin, alpha, NN in zip(xmin_arr, alpha_hats, Ncounts):
  axz[0].plot(
    xgrid, 
    power_law_density_fn(xgrid, alpha, xmin, N=NN),
    lw=0.75,
    color=xgrid_clrs.to_rgba(xmin),
  )
axz[0].set_ylim(ylim_)
fig.colorbar(xgrid_clrs, ax=axz[0])
```

For readmission, preliminary linear regression on log-log shows that we have $\alpha < 1$, so we cannot use the result from @clauset_power-law_2009. 
```{python}
#| label: fig-readmit-gridsearch 
#| fig-cap: Readmission time grid search 
readmit_df = pl.read_parquet(root / "data/concordant_networks/readmission_time.parquet")

readmit_xfx = pull_xfx(readmit_df, make_midpoint=False, quantity='readmission_time')

r_xmin_arr = np.arange(1, 200)
power_law_fits = [power_law_fit_graphical(*readmit_xfx, xmin=xm) for xm in r_xmin_arr]
pl_alphas, pl_icpts = zip(*power_law_fits)

fig, axs = plt.subplot_mosaic(
  [
    ['datafit', 'alpha'], 
    ['datafit', 'coeficpt'],
  ], 
  figsize=[10, 4],
)

axs['alpha'].plot(r_xmin_arr, pl_alphas)
# axs['alpha'].set_xlabel(r'$x_\text{min}$')
axs['alpha'].set_ylabel(r'$\hat\alpha$')

axs['coeficpt'].plot(r_xmin_arr, pl_icpts)
axs['coeficpt'].set_xlabel(r'$x_\text{min}$')
axs['coeficpt'].set_ylabel(r'$\hat{c}$')

axs['datafit'].scatter(*readmit_xfx, alpha=0.3, color='k', zorder=997)
axs['datafit'].set_xscale('log')
axs['datafit'].set_yscale('log')
axs['datafit'].set_xlabel('Readmission Duration [Days]')
axs['datafit'].set_ylabel('Count')
ylim_ = axs['datafit'].get_ylim()

r_xgrid = np.linspace(0.5, 370, 10_000)
r_xmin_norm = mcolors.Normalize(vmin=min(r_xmin_arr), vmax=max(r_xmin_arr))
r_xmin_clrs = cm.ScalarMappable(cmap=cm.YlOrRd, norm=r_xmin_norm)
for xmin, (alpha, icpt) in zip(r_xmin_arr, power_law_fits):
  axs['datafit'].plot(
    r_xgrid, 
    icpt * r_xgrid ** alpha,
    lw=0.75,
    color=r_xmin_clrs.to_rgba(xmin),
  )
axs['datafit'].set_ylim(ylim_)
fig.colorbar(r_xmin_clrs, ax=axs['datafit'])
```
